Separating Use and Reuse to Improve Both

### 2017-05-20 FACS/Intro.tex
.

\subsection{List of contribution}
\marco{text to improve, but is nice to keep a list}
\begin{itemize}
\item simple approach to fully separate inheritance and subtyping
\item easy to grasp model of compilation were typing is late but not
  too late (lazy?)\bruno{this idea is important: it's a key design
    decision. We need to talk about it in the introduction and offer
    some motivation/justification. }
\item clean and elegant handling of state in trat/module composition language
\item extension with nested classes is natural and
  powefull\bruno{wondering if this is necessary. I think there are
    several interesting/novel ideas already. Covering too much will
    result on not saying enough about each individual ideas.}
\item more operators can be accomodated without changing the general model
\end{itemize}
\input{section2}
\input{section3}
\input{section4}

\section{Old: Self Types, Binary Methods and the Need to Separate
  Inheritance from Subtyping}

So, far it appears that we have not gained much from decoupling
reuse from subtyping.

As discussed in Section~\ref{} the type of the self-reference is the
self-type. Like languages with ThisType, \name allows explicit
references to the self type. This is useful in several situations: 
For example it can be helpful in the presence of binary
methods~\cite{}, or methods that return an instance of the 
class being defined.

>>>>

\subsection{List of contribution}
\marco{text to improve, but is nice to keep a list}
\begin{itemize}
\item simple approach to fully separate inheritance and subtyping
\item easy to grasp model of compilation were typing is late but not
  too late (lazy?)\bruno{this idea is important: it's a key design
    decision. We need to talk about it in the introduction and offer
    some motivation/justification. }
\item clean and elegant handling of state in trat/module composition language
\item extension with nested classes is natural and
  powefull\bruno{wondering if this is necessary. I think there are
    several interesting/novel ideas already. Covering too much will
    result on not saying enough about each individual ideas.}
\item more operators can be accomodated without changing the general model
\end{itemize}
\input{section2}
\input{section4}
\input{section3}

### 2017-05-20 FACS/section4.tex
.

\begin{itemize}
\item keep sum associative and commutative,
\item allowing a class to create instances of itself,
\item actually initialize objects, leaving no null fields,
\item managing fields in a way that borrows the elegance of summing methods,
\item make it easy to add new fields.
\end{itemize}

@@

\begin{lstlisting}
  pointSum: { method int x()  method int y()//getters
    class method This of(int x,int y)//factory method
    method This sum(This other)
      This.of(this.x+other.x,this.y+other.y)//self instantiation
    }
  pointMul: { method int x() method int y()
    class method This of(int x,int y)
    method This mul(This other)
      This.of(this.x*other.x,this.y*other.y)
    }
\end{lstlisting}

@@

With our approach to handle the state, the following code can be directly composed:

\begin{lstlisting}
 //same code as before for pointSum and pointMul
  pointSum: { method int x() method int y()
    class method This of(int x,int y)
    method This sum(This other)
      This.of(this.x+other.x,this.y+other.y)
    }
  pointMul: { method int x() method int y()//look we are repeating
    class method This of(int x,int y)//the abstract method declarations.
    method This mul(This other)
      This.of(this.x*other.x,this.y*other.y)
    }
  PointAlgebra:Sum pointSum,pointMul
\end{lstlisting}

  As you can see, we can declare the methods independently and compose the result
  as we wish. However we have to repeat the abstract methods \Q@x@,\Q@y@ and \Q@of@.
  In addition of \Q@Sum@ and \Q@Mul@ we may want many operations over points; can we improve our reuse
  and not repeat such abstract definitions? of course!

\begin{lstlisting}
  p: { method int x() method int y()
    class method This of(int x,int y)
    }
  pointSum:Use p, { method This sum(This other)
      This.of(this.x+other.x,this.y+other.y)
    }
  pointMul:Use p, { method This mul(This other)
      This.of(this.x*other.x,this.y*other.y)
    }
  pointDiv: ...
  PointAlgebra:Use pointSum,pointMul,pointDiv,...
\end{lstlisting}

@@

What happens if we want to add fields instead of just operations?

\begin{lstlisting}
  colored:{ method Color color() }
  Point:Sum pointSum,colored //fails
\end{lstlisting}

@@

where we assume to support overloading on different parameter number.
This is a good solution, we think is better that any alternatives in literature,
however the method \Q@CPoint.sum@ resets the color to red.
What should be the behaviour in this case?
If we support withers, instead of writing \Q@This.of()@ we can use
\Q@this.withX(newX).withY(newY)@ in order to preserve the color from \Q@this@.
Sadly, if we use this design inside of \Q@sum(This other)@ we would loose the color from \Q@other@.

If the point designer could predict this kind of extension, then we could use the following design:  
\begin{lstlisting}
  p: { method int x() method int y()
    method This withX(int that)
    method This withY(int that)
    static method This of(int x,int y)
    method This merge(This other)
    }
  pointSum:Use p, { method This sum(This other)
      this.merge(other).withX(this.x+other.x).withY(this.y+other.y)
    }
  colored:{method Color color()
    method This withColor(Color that)
    method This merge(This other)
      this.withColor(this.color().mix(that.color())
    }
  CPoint:/*as before*/
\end{lstlisting}  
  Now we can merge colors, or any other kind of state we may want to add
  following this pattern.
  In order to compose, let say \Q@colored@ with \Q@flavored@ we would
  need to compose the merge operation inside of both of them.
  The simple model we are presenting could accomodate this with an
  extension allowing code literals inside of a \use\ expression to use some form of super call to compose conflicting implementations. This is similar to the \emph{override} operation present in the original trait model~\cite{ducasse2006traits}.

>>>>

\begin{itemize}
%complicated discussions on this point \item keep sum associative and commutative,
\item managing fields in a way that borrows the elegance of summing methods,
\item actually initialize objects, leaving no null fields,
\item make it easy to add new fields,
\item allowing a class to create instances of itself.
\end{itemize}

@@

\begin{lstlisting}
  pointSum: { method int x()  method int y()//getters
    class method This of(int x,int y)//factory method
    method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())//self instantiation
    }
  pointMul: { method int x() method int y()
    class method This of(int x,int y)
    method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
\end{lstlisting}

@@

\paragraph{The solution in \name:}${}_{}$\\
With our approach to handle the state, the following code can be directly composed:

\begin{lstlisting}
 //same code as before for pointSum and pointMul
  pointSum: { method int x() method int y()
    class method This of(int x,int y)
    method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())
    }
  pointMul: { method int x() method int y()//look we are repeating
    class method This of(int x,int y)//the abstract method declarations.
    method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
  PointAlgebra:Sum pointSum,pointMul
\end{lstlisting}

  As you can see, we can declare the methods independently and compose the result
  as we wish. In this first attempt, we repeated the abstract methods \Q@x@,\Q@y@ and \Q@of@.
  In addition of \Q@Sum@ and \Q@Mul@ we may want many operations over points; can we improve our reuse
  and not repeat such abstract definitions? of course!

\begin{lstlisting}
  p: { method int x() method int y()
    class method This of(int x,int y)
    }
  pointSum:Use p, { method This sum(This that)
      This.of(this.x()+that.x(),this.y()+that.y())
    }
  pointMul:Use p, { method This mul(This that)
      This.of(this.x()*that.x(),this.y()*that.y())
    }
  pointDiv: ...
  PointAlgebra:Use pointSum,pointMul,pointDiv,...
\end{lstlisting}

@@

\paragraph{Adding state:}${}_{}$\\
What happens if we want to add fields instead of just operations?

\begin{lstlisting}
  colored:{ method Color color() }
  Point:Use pointSum,colored //Failure: resulting class not coherent
\end{lstlisting}

@@

where we assume to support overloading on different parameter number.
This is a good solution, indeed we think is better that any alternatives in literature,
however the method \Q@CPoint.sum@ resets the color to red: we call the \Q@of(int,int)@ method, that now delegate to \Q@of(int,int,Color)@ by passing red as the default field value.
What should be the behaviour in this case?
If our abstract state supports withers, instead of writing \Q@This.of(...)@ we can use
\Q@this.withX(newX).withY(newY)@ in order to preserve the color from \Q@this@.
This solution is still not sadisfactory: this design ignore the color from \Q@that@.

If the point designer could predict this kind of extension, then we could use the following better design:  
\begin{lstlisting}
  p: { method int x() method int y() //getters
    method This withX(int that) method This withY(int that)//withers
    class method This of(int x,int y)
    method This merge(This that) //new method merge!
    }
  pointSum:Use p, { method This sum(This that)
      this.merge(that).withX(this.x()+that.x()).withY(this.y()+that.y())
    }
  colored:{method Color color()
    method This withColor(Color that)
    method This merge(This that) //how to merge color handled here
      this.withColor(this.color().mix(that.color())
    }
  CPoint:/*as before*/
\end{lstlisting}  
  Now we can merge colors, or any other kind of state we may want to add
  following this pattern.

  In order to compose, let say \Q@colored@ with \Q@flavored@ we would
  need to compose the merge operation inside of both of them.
  The simple model we are presenting could accomodate this with an
  extension allowing code literals inside of a \use\ expression to use some form of super call to compose conflicting implementations. This is similar to the \emph{override} operation present in the original trait model~\cite{ducasse2006traits}.

In the following you can see how to mix colors and flavours. The syntax \Q@_2merge@ and   \Q@_3merge@
call the version of merge as defined in the second/third element of \use.
\begin{lstlisting}
  p: {/*as before*/ }
  pointSum:Use p, {/*as before*/ }
  colored:{/*as before*/}
  flavored:{method Flavor flavor() //very similar to colored
    method This withFlavor(Flavor that)
    method This merge(This that) //how to merge flavors handled here
      this.withFlavor(that.flavor())}//here we just inherit ``that'' flavor
  FCPoint:Use pointSum,colored,flavored{
    class method This of(int x,int y)
      This.of(x,y,Color.of(/*red*/),Flavor.none())
    class method This of(int x, int y,Color color,Flavor flavor)
    //we can resolve the conflict about two implementations for merge
    //by proving our own implementation here
    method This merge(This that) this._2merge(that)._3merge(that)
    }
\end{lstlisting}
